Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-05-31 writeup/pnas-word-order-si.tex

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{tikz-dependency}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\E}{\mathop{\mathbb{E}}}

@@

@@

\section{Results on all UD Relations}
Table~\ref{tab:all-predictions} shows the predicted prevalence of correlations between the \emph{obj} dependency and all UD dependency types, along with the expected prevalence according to typological studies.

\begin{table*} % TODO figure out away to make this table more aesthetic
	\begin{center}

\tiny{
\begin{tabular}{|l|l|l|ll|l|l|}
	\hline
	Relation & Real & DepL & Pred & Pars & Efficiency & Expected Prevalence  \\ \hline
% From Dryer	
acl  &  80   &   \textbf{85}$^{***}$   &   43   &   \textbf{82}$^{***}$   &   \textbf{71}$^{***}$  & $>50\%$  \citep{dryer1992greenbergian} \\
aux  &  12   &   \textbf{26}$^{***}$   &   \textbf{19}$^{***}$   &   49   &   \textbf{35}$^{**}$  & $<50\%$ \citep{dryer1992greenbergian} \\
lifted\_case  &  86   &   \textbf{81}$^{***}$   &   \textbf{40}$^{**}$   &   \textbf{77}$^{***}$   &   \textbf{67}$^{***}$ & $> 50 \%$ \citep{dryer1992greenbergian}  \\
lifted\_cop  &  94   &   \textbf{80}$^{***}$   &   59   &   \textbf{72}$^{***}$   &   \textbf{63}$^{**}$  & $> 50 \%$ \citep{dryer1992greenbergian} \\
lifted\_mark  &  76   &   \textbf{85}$^{***}$   &   \textbf{57}$^{*}$   &   \textbf{76}$^{***}$   &   \textbf{72}$^{***}$  & $> 50 \%$ \citep{dryer1992greenbergian} \\
nmod  &  80   &   \textbf{82}$^{***}$   &   49   &   \textbf{71}$^{***}$   &   \textbf{68}$^{***}$  & $> 50 \%$ \citep{dryer1992greenbergian} \\ 
obl  &  88   &   \textbf{78}$^{***}$   &   \textbf{71}$^{***}$   &   46   &   \textbf{68}$^{***}$  &  $> 50 \%$ \citep{dryer1992greenbergian} \\
xcomp  &  88   &   \textbf{90}$^{***}$   &   \textbf{76}$^{***}$   &   \textbf{89}$^{***}$   &   \textbf{85}$^{***}$  &  $> 50 \%$ \citep{dryer1992greenbergian} \\
\hdashline
nsubj  &  33   &   \textbf{29}$^{***}$   &   53   &   \textbf{7}$^{***}$   &   \textbf{23}$^{***}$  & $> 50 \%$ \citep{dryer1992greenbergian} \\
advmod  &  35   &   51   &   \textbf{18}$^{***}$   &   47   &   \textbf{39}$^{**}$  & $>50\%$ for manner adverbs; $\approx 50\%$ for negation, TAM, intensifiers \citep{dryer1992greenbergian} \\

\hline
% Other universals, not in Dryer
advcl  &  86   &   \textbf{84}$^{***}$   &   51   &   \textbf{67}$^{***}$   &   \textbf{67}$^{*}$  & $>50\%$ \citep{greenberg1963universals,diessel2001ordering} \\
amod  &  51   &   \textbf{84}$^{***}$   &   42   &   \textbf{53}$^{*}$   &   51   & $\approx 50\%$ \citep{dryer1992greenbergian} \\
nummod  &  37   &   \textbf{74}$^{**}$   &   49   &   52   &   52 & $\approx 50\%$ \citep[][89A vs. 83A]{wals} \\

\hline
% Typological data not available
appos  &  77   &   \textbf{77}$^{***}$   &   49   &   \textbf{67}$^{***}$   &   \textbf{62}$^{**}$ &  Unknown \\%($>50\%$? correlates with \textit{nmod} according to Greenberg U 23)  \\ % \cite{cinque2009greenberg} claims it correlates with OV/VO. However, it's not clear to me how closely appos matches the subject of the typological claim
lifted\_cc  &  71   &   \textbf{83}$^{***}$   &   51   &   \textbf{81}$^{***}$   &   \textbf{71}$^{***}$  & Unknown \\
ccomp  &  86   &   \textbf{88}$^{***}$   &   \textbf{70}$^{**}$   &   \textbf{77}$^{***}$   &   \textbf{77}$^{**}$ & Unknown (cf. \cite{dryer1980positional}) \\ % TODO According to Dryer 1980, there is a correlation
csubj  &  79   &   \textbf{78}$^{***}$   &   60   &   \textbf{66}$^{***}$   &   \textbf{68}$^{**}$  & Unknown (cf. \cite{dryer1980positional}) \\% TODO According to Dryer 1980, there is a correlation
expl  &  12   &   36   &   62   &   40   &   53  & Unknown \\
iobj  &  74   &   49   &   \textbf{78}$^{***}$   &   \textbf{38}$^{*}$   &   62  & Unknown \\
vocative  &  46   &   \textbf{42}$^{*}$   &   59   &   45   &   \textbf{45}$^{*}$  & Unknown \\

\hline
% Uninterpretable
compound  &  66   &   \textbf{61}$^{*}$   &   48   &   \textbf{58}$^{*}$   &   52  & Uninterpretable \\
det  &  31   &   \textbf{71}$^{***}$   &   \textbf{56}$^{*}$   &   48   &   54  & Uninterpretable (conflates articles and others) \\%$.50\%$ for articles, $\approx 50\%$ for others, \citep{dryer1992greenbergian} \\ %TODO what else is conflated in det? why do we keep advmod but not det?
dislocated  &  47   &   61   &   \textbf{68}$^{**}$   &   55   &   58  & Uninterpretable \\
dep  &  56   &   \textbf{64}$^{*}$   &   52   &   58   &   \textbf{58}$^{*}$   & Uninterpretable \\

\end{center}
\caption{Predictions on all UD relations occurring in more than one language. Numbers are the fraction of languages where the direction of the given relation agrees in direction with the \emph{obj} relation. In the last column, we indicate what direction would be expected typologically. ``Uninterpretable'' UD relations are those which collapse so many different linguistic relationships that they are not linguistically meaningful. ``UD artifact'' relations are those whose order is determined strictly by UD parsing standards, such that their order is not linguistically meaningful: these include dependencies such as the connection between two parts of a word that have been separated by whitespace inserted as a typo (\emph{goeswith}).}
\label{tab:all-predictions}
\end{table*}

\section{Results on Best Optimized Grammars}

In the main part, we reported the percentage of optimized grammars supporting correlations.
Here we report these numbers for the grammars that optimize each objective most successfully.
That is, for each language and objective, we chose among the eight optimized grammars the one that achieves the best value for the objective function on the validation partition of the treebanks -- that is, the one where the optimization procedure was most successful.

Numbers for these languages are shown in Table~\ref{table:results-dryer-best-only}.
%Confirming the results from the main paper, 
As expected, the correlations mostly have stronger prevalences than when considering all optimized languages.
For Dependency Length, the correlations that it predicts mostly have prevalences close to 100 \%, whereas both the UD languages and efficiency-optimized languages show somewhat lower percentages.
That is, dependency length mostly overshoots for the correlations that it predicts correctly.

\begin{table*}
	\begin{center}
\begin{tabular}{|ll|l|l|l|ll|l|}
	\hline
	%&
	\multicolumn{2}{|c|}{Correlates with...}   &          \multirow{2}{*}{Operationalization}       &     \multirow{2}{*}{Real}   & \multirow{2}{*}{DepL}  & \multirow{2}{*}{Pred}   &  \multirow{2}{*}{Pars} &  \multirow{2}{*}{Efficiency}  \\ 
	verb & object &      & & & &  & \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
adposition    &    NP    &    lifted\_case    &    86    &    98    &    69    &    88    &    83   \\
copula    &    NP    &    lifted\_cop    &    94    &    98    &    58    &    85    &    74   \\
auxiliary    &    VP    &    aux    &    88    &    84    &    90    &    61    &    73   \\
noun    &    genitive    &    nmod    &    80    &    98    &    75    &    88    &    91   \\
noun    &    relative clause    &    acl    &    80    &    100    &    69    &    88    &    90   \\
complementizer    &    S    &    lifted\_mark    &    76    &    100    &    76    &    82    &    87   \\
verb    &    PP    &    obl    &    88    &    98    &    88    &    75    &    91   \\
want    &    VP    &    xcomp    &    88    &    100    &    92    &    94    &    96   \\
verb    &    subject    &    nsubj    &    33    &    25    &    53    &    10    &    21   \\
verb    &    manner adverb    &    advmod    &    35    &    61    &    27    &    55    &    36   \\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
\end{tabular}
	\end{center}
	\caption{Percentages of grammars supporting correlations, among grammars that best optimize each objective on treebanks of a given language.}\label{table:results-dryer-best-only}
\end{table*}

\begin{figure}
    \centering

    \includegraphics[scale=.25]{../results/correlations/figures/coverage-langmod-best.pdf}
    \includegraphics[scale=.25]{../results/correlations/figures/coverage-parse-best.pdf}
    \includegraphics[scale=.25]{../results/correlations/figures/coverage-two-lambda09-best.pdf}

    \includegraphics[scale=.25]{../results/correlations/figures/coverage-ground.pdf}
    \includegraphics[scale=.25]{../results/correlations/figures/coverage-depl.pdf}
	\caption{Correlations satisfied by the best word order grammars for predictability, parseability, efficiency, by the UD languages, and by the best grammars for dependency length.}
    \label{fig:posterior-prevalence}
\end{figure}

\section{Further Statistical Analyses}

In the main results table, we reported the fraction of optimized languages satisfying a correlation.
We show that essentially the same predicted fractions are obtained when controlling for the imbalance in distribution of language families.

To this end, we used the logistic mixed-effects regressions that were used to estimate significance to compute the expected prevalence of a correlation controlling for language families.
To obtain predictions controlling for languages and language families, we plot the logit-transformed posterior of $\alpha$ in Figure \ref{fig:posterior-prevalence}.

\begin{figure}
    \centering
    \includegraphics[scale=.7]{../results/correlations/figures/posterior_joint_with_ud.pdf}
	\caption{Posteriors of the prevalences of the correlations among languages optimized for Predictability (red), parseability (green), and efficiency (blue), controlling for languages and language families, with prevalence in the UD languages.}
    \label{fig:posterior-prevalence}
\end{figure}

%\begin{figure}
%    \centering
%    \includegraphics[scale=.3]{figures/posterior-satisfied-universals-efficiency.pdf}
%     \includegraphics[scale=.3]{figures/posterior-satisfied-universals-langmod.pdf}
%     \includegraphics[scale=.3]{figures/posterior-satisfied-universals-parseability.pdf}
%    \includegraphics[scale=.3]{figures/posterior-satisfied-universals-depl.pdf}
%	\caption{Posterior of the number of correlations predicted in the direction found in the UD treebanks.}
%    \label{fig:posterior-satisfied-count}
%\end{figure}

%We argued that efficiency predicts all 10 correlation universals in the direction manifested in the UD corpora, as for each universal, the posterior probability that less than $50 \%$ of the optimized grammars satisfy the correlation was less than $0.05$.
%To address potential multiple comparisons problems, we use a single logistic model to show that efficiency predicts all ten correlations with high posterior probability.

%For each objective function, we constructed a single logistic model predicting, for each of the ten dependencies, whether it is correlated or anti-correlated with the \emph{obj} dependency, with random effects for language and language family, correlated across the ten dependencies.

%We used this model to estimate the posterior distribution of the number of correlations that an objective function predicts to be in the same direction as found in the UD treebanks.
%The resulting distributions are shown in Figure~\ref{fig:posterior-satisfied-count}.
%The estimated posterior probability that efficiency predicts less than all ten dependencies to correlate in the same direction as in the UD treebanks is $0.0242$. The probability that it predicts less than nine of the correlations is $3\cdot 10^{-4}$.

\section{Relation to the Information Bottleneck}

Our efficiency objective (Equation~\ref{eq:efficiency}), to be maximized for word order grammars $L$, is restated below.
\begin{equation}
    \label{eq:efficiency-again}
    J_T(L) = I[L(T); T] - \lambda H[L(T)].
\end{equation}
This equation can be seen as a special case of the objective function proposed in \citet{tishby1999information} for general lossy compression, based on rate-distortion theory \citep{cover2006elements,harremoes2007information}. In that theory, the goal is to derive an encoding $\hat{X}=L(X)$ of some variable $X$ which preserves the information in $X$ relevant to some third variable $Y$, while minimizing irrelevant information. The three variables $Y$,$X$, and $\hat{X}$ are assumed to form a Markov chain $Y \rightarrow X \underrightarrow{L} \hat{X}$. Given those variables, the information bottleneck objective to be maximized is given in Eq.~\ref{eq:ib}.
\begin{equation}
    \label{eq:ib}
    J_{X,Y}(L) = I[\hat{X}; Y] - \lambda I[\hat{X}; X].
\end{equation}
In this equation, the joint distribution of $X$ and $Y$ is seen as fixed, and the conditional distribution $L=\hat{X}|X$ is optimized.

In our setting we consider a language $L$ expressing trees $T$ as utterances $U$: $T \underrightarrow{L} U$. To make our formulation more parallel to the information bottleneck, we assume the utterances $U$ are passed through a channel from producer to comprehender, and received as $\hat{U}=U$. We have the Markov chain $T \underrightarrow{L} U \rightarrow \hat{U}$, where we are assuming the conditional channel distribution $\hat{U}|U$ is fixed and the conditional distribution $L=U|T$ is to be optimized. Then our objective is parallel to Eq.~\ref{eq:ib}:
\begin{equation}
    \label{eq:efficiency-kinda-ib}
    J_T(L) = I[U;T] - \lambda I[U;\hat{U}]. % TODO should this be I[\hat{U};T]?
\end{equation}
Assuming the channel $U \rightarrow \hat{U}$ is noiseless, we have $U=\hat{U}$, and thus $I[U;\hat{U}]=I[U;U] = H[U]$. Then Eq.~\ref{eq:efficiency-kinda-ib} reduces to a form of Eq.~\ref{eq:efficiency-again}:
\begin{equation*}
    J_T(L) = I[U;T] - \lambda H[U].
\end{equation*}

Our efficiency objective is thus closely related to the information bottleneck; the difference is that we are optimizing the mapping from underlying meanings to representations (strings) assuming that these strings will be transmitted in a fixed noiseless channel.

\section{Possible values of $\lambda$}

In the efficiency objective~(\ref{eq:efficiency-again}), the value of $\lambda$ is constrained to be in $[0,1)$.
This means, surprisal must be weighted less strongly than parsability.

Greater values of $\lambda$ mathematically result in degenerate solutions.
To show this, note that the following inequality always holds:
\begin{equation}
I[L(T); T] \leq H[L(T)].
\end{equation}
Therefore, if $\lambda \geq 1$, the efficiency objective satisfies 
\begin{equation}
    J_T(L) = I[L(T); T] - \lambda H[L(T)] \leq 0.
\end{equation}
and it takes the maximally possible value zero if $L(T)$ maps all trees to a single utterance, in which case both $I[L(T); T]$ and $H[L(T)]$ are zero.
This is a degenerate language with only a single utterance, which is simultaneously used to convey all meanings.
While the design of our word order grammars precludes a collapse of all trees to a single utterance, this shows that an objective with $\lambda \geq 1$ cannot be a generally applicable objective for language efficiency.

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parsability.
We previously carried out language optimization for $\lambda = 1$, and obtained essentially identical predictions for word order     correlations and relative optimality of language.

\section{Joint optimization of grammar, parser, and language model parameters}

For each of our objective functions, we seek parameters $\theta$ of a word order grammar $L_\theta$ that maximize the average utility of sentences in the language obtained by linearizing unordered trees $t \sim T$ according to $L_\theta$. The calculation of the predictability and parseability objectives also requires fitting language models parameterized by $\phi$ and parsers parameterized by $\psi$, respectively; and the calculation of the efficiency objective requires joint optimization of the word order grammar, language models, and parsers. The optimization problems come out to:
\begin{align}
\max_\theta &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{DepL}(t; {\bf w}) \label{eq:maximize-depl}\\
\max_{\theta,\phi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pred}^{\phi(\theta)}(t; {\bf w}) \label{eq:maximize-pred}\\
\max_{\theta,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pars}^{\psi(\theta)}(t; {\bf w}) \label{Eq:maximize-pars} \\
\max_{\theta,\phi,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Efficiency}^{\psi(\theta), \phi(\theta)}(t; {\bf w}). \label{eq:maximize-efficiency} 
\end{align}
Note that the original word orders from the treebank never enter this objective---in particular, they do not influence the language model and the parser, which are entirely determined by $\theta$.

The case of dependency length in Eq.~\ref{maximize-depl} is the simplest, requiring only coordinate descent on the parameters $a_\tau, b_\tau \in \theta$. A similar optimization problem was solved by hill-climbing in \citet{gildea2007optimizing} and \citet{gildea2015human}.
We optimize this objective by stochastic gradient descent with a batch size of one unordered dependency tree.

>>>>

\documentclass[10pt,twoside,lineno]{article}
% Use the lineno option to display guide line numbers if required.

%\usepackage{lineno}
%\linenumbers

%\templatetype{pnassupportinginfo}

% \readytosubmit %% Uncomment this line before submitting, so that the instruction page is removed.

\usepackage{amsmath}
\usepackage{tikz-dependency}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\E}{\mathop{\mathbb{E}}}

\usepackage{footnote}
\makesavenoteenv{tabular}
\makesavenoteenv{table}

\usepackage{siunitx}

\usepackage[cm]{fullpage}

\usepackage{longtable}

\usepackage[numbers]{natbib}

@@

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thesection}{S\arabic{section}}

\newcommand{\utterance}{\mathcal{U}}
\newcommand{\tree}{\mathcal{T}}

@@

\tableofcontents

\textcolor{red}{run czech parser}

\textcolor{red}{keep in mind: pareto plot unify with SI plot}

\section{Formalization of Greenberg Correlation Universals}\label{sec:correlations}

Here we describe how we selected the word order correlations in Table 1 of the main paper, and how we formalized these using syntactic relations defined by Universal Dependencies.

We base our formalization on the comprehensive study by Dryer \cite{dryer1992greenbergian}.\footnote{Regarding objections by \citet{dunn2011evolved}, we refer to the follow-ups by \citet{levy2011computational}, and \citet{croft2011greenbergian}.}
Greenberg's original study was based on 30 languages; more recently, Dryer \cite{dryer1992greenbergian} documented the word order correlations based on typological data from 625 languages.
\cite{dryer1992greenbergian} formulated these universals as correlations between the order of objects and verbs and the orders of other syntactic relations.
We test our ordering grammars for these correlations by testing whether the coefficients for these syntactic relations have the same sign as the coefficient of the verb-object relation.
Testing correlations is therefore constrained by the degree to which these relations are annotated in UD.
The verb-object relation corresponds to the  \emph{obj} relation defined by UD.
While most of the other relations also correspond to UD relations, some are not annotated reliably.
We were able formalize eleven out of Dryer's seventeen correlations in UD.
Six of these could not be expressed individually in UD, and were collapsed into three coarse-grained correlations:
First, tense/aspect and negative auxiliaries are together represented by the \emph{aux} relation in UD.
Second, the relation between complementizers and adverbial subordinators with their complement clauses is represented by the \emph{mark} relation.
Third, both the verb-PP relation and the relation between adjectives and their standard of comparison is captured by the \emph{obl} relation.

The resulting operationalization is shown in Table~\ref{table:greenberg-dryer}.
For each relation, we show the direction of the UD syntactic relation: $\rightarrow$ indicates that the verb patterner is the head; $\leftarrow$ indicates that the object patterner is the head.

As described in \emph{Materials and Methods}, we follow \citet{futrell2015largescale} in converting the Universal Dependencies format to a format closer to standard syntactic theory, promoting adpositions, copulas, and complementizers to heads.
As a consequence, the direction of the relations \emph{case}, \emph{cop}, and \emph{mark} is reversed compared to Universal Dependencies.
For clarity, we refer to these reversed relations as \emph{lifted\_case}, \emph{lifted\_cop}, and \emph{lifted\_mark}.

\begin{table*}[ht]
	\begin{center}
\small{
\begin{tabular}{|l|ll|l|l|}
	\hline
&	\multicolumn{2}{c|}{Correlates with...}   &          \multirow{2}{*}{UD Relation}   & \multirow{2}{*}{\citet{greenberg1963universals}}    \\ 
&	verb & object & &   \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {1}}}&adposition    &    NP    &  $\xrightarrow{lifted\_case}$   & 3, 4   \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {2}}}&copula  verb  &    predicate    &    $\xrightarrow{lifted\_cop}$   & --    \\\hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}}&tense/aspect auxiliary    &    VP    &    \multirow{2}{*}{$\xleftarrow{aux}$}   & 16, 13  \\
&	negative auxiliary    &    VP    &    & -- \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}}&noun    &    genitive    &   $\xrightarrow{nmod}$ & 2, 23   \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {5}}}&noun    &    relative clause    &    $\xrightarrow{acl}$ &  24    \\ \hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {6}}}}&complementizer    &    S    &   \multirow{2}{*}{$\xrightarrow{lifted\_mark}$}  & --    \\
&	adverbial subordinator & S &  & -- \\ \hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}}&	adjective & std. of comp. & \multirow{2}{*}{$\xrightarrow{obl}$} & --\\
&verb    &    PP    &    & 22   \\\hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}&`want'    &    VP    &    $\xrightarrow{xcomp}$   & 15   \\\hline
\end{tabular}
}
	\end{center}
	\caption{Greenbergian Correlations based on Dryer \cite{dryer1992greenbergian}, with operationalizations with Universal Dependencies using the modified format of \cite{futrell2015largescale} (see text).
	For reference, we also provide the numbers of the closest corresponding universals stated in Greenberg's original study, to the extent that this is possible.
	}\label{table:greenberg-dryer}
\end{table*}

\paragraph{Excluded Correlations}
We excluded three correlations that are not annotated reliably in UD, and are only relevant to some of the world's languages: Question particles, plural words (i.e., independent plural markers), and articles.
All three types of elements occur at most in parts of the 51 UD languages, and none of them is annotated reliably in those languages where they occur.
Among these three types of elements, the one most prominent in our sample of 51 languages is articles (occurring in many European languages); UD subsumes them under the \emph{det} relation, which is also used for other frequent elements, such as demonstratives and quantifiers.

We also excluded the verb-manner adverb correlation.
UD does not distinguish manner adverbs from other elements labeled as adverbs, such as sentence-level adverbs and negation markers, whose ordering is very different from manner adverbs.
All types of adverbs are unified under the \emph{advmod} relation.
In the real orderings in our sample of 51 UD languages, the dominant ordering of \emph{advmod} almost always matches that of subjects -- that is, \emph{advmod} dependents are ordered after the verb only in VSO languages.
This ordering behavior is very different from that documented for manner adverbs by Dryer.

We further excluded the verb-subject correlation, which is not satisfied by much more than half of the world's languages (51 \% among those with annotation in the \emph{World Atlas of Language Structures} \cite{wals-81}, with clear violation in 35.4 \%).
It is satisfied only in 33\% of our sample of 51 UD languages, as quantified using the grammars we extracted.
Dryer \cite{dryer1992greenbergian} counts this as a correlation since he describes the distribution of subject order as an interaction between a weak correlation with object order, and a very strong dominance principle favoring SV orderings.
We focus on the modeling of correlations, and leave dominance principles to future research.
We therefore excluded this correlation here.
%We note that efficiency makes predictions compatible with the pattern observed in our sample of 51 languages, potentially in line with the noisy-channel explanation of SVO described by~\cite{gibson2013noisy} (see Table~\ref{tab:all-predictions}).

%croft2011greenbergian
%dryer2011evidence

\paragraph{Other Greenberg Universals}
Greenberg~\cite{greenberg1963universals} stated a total of 45 universals.
Twenty of these concern the structure of individual words (as opposed to word order, which we focus on here), and many of those have been argued to be explained by the ``dual pressures'' idea \cite{haspelmath2006against}.
The other 25 universals concern word order; Dryer \cite{dryer1992greenbergian} reformulated most of these as correlations with verb-object order; these form the basis of our formalization in Table~\ref{table:greenberg-dryer}. %, finding 16 correlations to likely be universal.
%Excluding universals not annotated in available databases, and merging some that can only be tested together we arrived at the eight correlations shown in Table~\ref{table:greenberg-dryer}.
%Testing efficiency on the excluded correlations would be feasible with additional annotation of data.
%For instance, one of these could be tested by labeling the existing databases for the distinction between manner adverbs from other kinds of adverbs.
%For another one, it would be necessary to annotate databases for additional languages that have independent plural markers (e.g., Hawaiian).
There are a few other well-supported word order universals that are not correlations with verb-object order.
This includes dominance principles~\cite{croft2003typology} such as the strong preference for subjects to precede objects.
Furthermore, there has been interest in Greenberg's universals 18 and 20, which describe correlations not with verb-object order, but of different elements of noun phrases~\cite{cinque2005deriving, culbertson2014language, dryer2018order}.
Future work should examine whether these universals can also be linked to efficiency optimization.

\section{Formalizing Comunicative Efficiency}
\subsection{Relation to Other Work}

%The efficiency of any information-theoretic communication protocol depends on (1) how costly encoding and transmission are, and (2) how precisely messages can be recovered from codes~\cite{shannon1948mathematical}.
%Accordingly, computational tests of functional explanations in linguistics have formalized the overall efficiency of language as a weighted combination of terms representing the amount of information that utterances contain about the underlying messages, and the cost of communication.

Here we discuss how our formalization of communicative efficiency relates to formalizations that have been proposed in the information-theoretic literature on language.
Across the literature, the core idea is to maximize the \key{amount of information} that linguistic forms provide about meanings, while constraining \key{complexity and diversity} of forms, with some differences in the precise formalization of these two quantities \cite{ferreri2003least,ferrericancho2007global, kemp2012kinship,xu2014numeral, regier2015word,xu2016historical, zaslavsky2018efficient,futrell2017memory,peloquin2019interactions, zaslavsky2019semantic, frank2012predicting,goodman2013knowledge, kao2014nonliteral, bennett2018extremely}.

The basis for our precise formalization is the function proposed in \cite{ferreri2003least,ferrericancho2007global,futrell2017memory,peloquin2019interactions} as a general efficiency metric for communicative systems.
%The key idea is to maximize the \key{amount of information} that linguistic forms provide about meanings, while constraining \key{complexity and diversity} of forms. %; there are some differences in the formalization of complexity.
If $S$ denotes signals (e.g., words, sentences) and $R$ denotes their referents (e.g., objects in a reference game), then this efficiency metric takes the form (notation slightly varies across these publications):
\begin{equation}\label{eq:eff-general}
I[S, R] - \lambda H[S]
\end{equation}
where $I[S, R]$ describes the \key{informativity} of the signals $S$ about their referents $R$, and $H[S]$ describes the \key{complexity} of the communication system.
While prior studies \cite{ferreri2003least, kemp2012kinship,regier2015word,zaslavsky2018efficient} mostly considered settings where the signals $S$ are individual words without further structure, the signals are entire sentences $\utterance$ in our setting.
The underlying messages $R$ which the speaker aims to convey are the syntactic structures $\tree$.
By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined.
The syntactic structure specifies how the meanings of words are combined; therefore, recovering the syntactic structure is a prerequisite to understanding a sentence correctly.
%The idea that efficient languages will permit the listener to infer grammati

We therefore arrive at the following efficiency metric:
\begin{equation}\label{eq:efficiency-derived}
	R_{\textit{Eff}} := R_{\textit{Parseability}} + \lambda R_\textit{Pred}
\end{equation}
where \key{parseability} is the amount of information that utterances provide about their underlying syntactic structures:
\begin{equation}
	R_{Pars} := I[\utterance,\tree] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
and \key{predictability} is the negative entropy or surprisal of the language:
\begin{equation}
	R_{Pred} := - H[\utterance] = \sum_{u} p(u) \log p(u)
\end{equation}

We now discuss how this metric relates to other variants that have been proposed in the information-theoretic literature on language, and describe why~(\ref{eq:eff-general}-\ref{eq:efficiency-derived}) is particularly appropriate to our setting.
This efficiency metric~(\ref{eq:eff-general}-\ref{eq:efficiency-derived}) is equivalent to a deterministic version \cite{strouse2017deterministic} of the Information Bottleneck approach, which has been successfully applied to modeling word meaning across different domains \cite{zaslavsky2018efficient, zaslavsky2019semantic}.
The Information Bottleneck models complexity using a mutual information term instead of the entropy appearing in~(\ref{eq:eff-general}).
This mutual information formalization is appropriate to codes that are nondeterministic \cite{zaslavsky2018efficient}.
As we assume deterministic grammars that transduce every underlying syntactic structure into one surface order, our model corresponds to the Deterministic Information Bottleneck~\cite{strouse2017deterministic}, which arises when the encoder is deterministic, and which uses the entropy as the complexity measure.
Furthermore, entropy also arises naturally as a complexity measure from the role played by surprisal in determining human language processing effort~\cite{hale2001probabilistic,levy2008expectation,smith2013effect}.

A few other approaches share the mutual information term for informativity in (\ref{eq:eff-general}), while using complexity measures that are not explicitly information-theoretic.
%A few models have used other complexity measures that are not explicitly information-theoretic quantities.
In \cite{regier2007color, xu2014numeral, xu2016historical}, the complexity function is the number of different forms; in \cite{kemp2012kinship} it is the complexity of defining the concepts that are encoded.
The number of forms is not applicable in our case, as languages will typically have infinitely many sentences.
Notably, the models in \cite{regier2007color, xu2016historical} have since been reformulated successfully in the Information Bottleneck formalism \cite{zaslavsky2018efficient, zaslavsky2019semantic}, bringing them even closer to our formalization of efficiency.

In addition to these models, which quantify the efficiency of communication systems, there is closely related work formalizing the optimal choice of specific utterances in context, such as models of pragmatic reasoning in reference games \cite{frank2012predicting,goodman2013knowledge, kao2014nonliteral}.
In line with the other models, these assume that rational speakers choose utterances to optimize informativity about the referent object, and trade this off with the cost of the utterance, which is partly chosen to be the surprisal of the utterance \cite{bennett2018extremely}.

%\textcolor{red}{also mention \cite{piantadosi2012communicative,fedzechkina2012language}}

%Other metrics very similar to~(\ref{eq:eff-general}) have been proposed in other computational work formalizing Zipf's ideas~\cite{frank2012predicting,zaslavsky2018efficient,kemp2012kinship,regier2015word,goodman2013knowledge,piantadosi2012communicative}.
%The idea of formalizing informativity using the mutual information $I[S, R]$ is common to all these approaches.

%These information-theoretic accounts vary somewhat in the formalizations of complexity.
%Like our model, \cite{ferreri2003least,ferrericancho2007global,futrell2017memory} model the complexity as the \key{entropy} of the distribution of forms, which .
%This model is closely related to the Information Bottleneck approach \cite{zaslavsky2018efficient, zaslavsky2019semantic}, which models complexity as the mutual information between forms (e.g., words) and their referents (e.g., individual objects).

\subsection{Possible values of $\lambda$}\label{sec:lambda}

In the efficiency objective
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Pars}} + \lambda R_\textit{Pred}
\end{equation}
the value of $\lambda$ is constrained to be in $[0,1)$.
This means, surprisal must be weighted less strongly than parseability.

Greater values of $\lambda$ can mathematically result in degenerate solutions.
To show this, note that the following inequality always holds:
\begin{equation}
I[\utterance; \tree] \leq H[\utterance].
\end{equation}
Therefore, if $\lambda \geq 1$, the efficiency objective satisfies 
\begin{equation}
    R_{Eff} = I[\utterance; \tree] - \lambda H[\utterance] \leq 0.
\end{equation}
and it takes the maximally possible value zero if there is only a single utterance $\utterance$, in which case both $I[\utterance; \tree]$ and $H[\utterance]$ are zero.
This is a degenerate language with only a single utterance, which is simultaneously used to convey all meanings.
While the design of our word order grammars (see \emph{Materials and Methods}) precludes a collapse of all syntactic structures to a single utterance, this shows that an objective with $\lambda \geq 1$ cannot be a generally applicable description of the efficiency of communication systems.
In conclusion, $\lambda$ is constrained to be in $[0,1)$, with values closer to $1$ placing similar weights on both predictability and parseability, whereas values closer to $0$ diminish the role of predictability.

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parseability.
While the computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$, we also examined word order predictions for grammars optimized for only parseability or only predictability, in order to tease apart predictions made by these two components.
As shown in Table~\ref{tab:all-predictions-1}, each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions.
%Indeed, the same holds for all of the 38 syntactic relations in UD  (Table~\ref{tab:all-predictions-1}-\ref{tab:all-predictions-2})
%We additionally report results with $\lambda = 0$, i.e., placing all weight on parseability.
That is, at $\lambda$ close to its maximal value, the predictions of optimizing the two scoring functions individually add up to the predictions of efficiency optimization.\footnote{Results from one of the preliminary experiments reported in Figure~\ref{table:corr-resu-previous} show that results are stable to small variation of $\lambda$: Essentially equivalent predictions are obtained for $\lambda=1.0$. While $\lambda=1.0$ is not a valid choice for communicative efficiency in general due to the possibility of collapse to a single utterance, our setting does not allow such a collapse, as the syntactic structure already determines which words are present in the sentence.}
Small values of $\lambda$ correspond to the case where predictability plays no role, and only parseability is optimized (Table~\ref{tab:all-predictions-1}), in which case not all correlations are predicted (Figure~\ref{fig:posterior}).
%Finally, the robustness of our results to small variations of $\lambda$ is confime by reporting results with $\lambda=1$ in Table~\ref{table:corr-resu-previous}.
